Manuscript for a Data Analysis Project: Diabetes Diagnosis

### 2024-07-27 products/manuscript/Manuscript.qmd
Completed Manuscript
Finalized the KNN model and then compared all three. Outlined the results and provided a summary as well as limitations. Once analysis was complete, wrote an abstract and completed readme files.

---
title: "Manuscript for a Data Analysis Project: Diabetes Diagnosis"
subtitle: ""
author: Austin Thrash
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    highlight-style: github
csl: ../../assets/apa.csl
---

@@

*\*Summary will be finalized in later part of project and will include setup and results*

@@

*\*more background information will be added later, I will describe what diabetes is, what are the notable influences on diagnosis, and what the dataset provides in relation to other studies on diabetes diagnosis.*

@@

|--------------------|----------------------------------------------------|

@@

@@

\*Will fix and put into readable table

```         
            HighBP             HighChol            CholCheck                  BMI               Smoker               Stroke HeartDiseaseorAttack 
            1.141777             1.070725             1.011236             1.124242             1.073702             1.071638             1.150480 
        PhysActivity               Fruits              Veggies    HvyAlcoholConsump        AnyHealthcare          NoDocbcCost              GenHlth 
            1.140832             1.109908             1.105040             1.013591             1.100212             1.157580             1.743109 
            MentHlth             PhysHlth             DiffWalk                  Sex                  Age            Education               Income 
            1.281663             1.697022             1.499052             1.109269             1.326329             1.299898             1.505502 
```

@@

***\*IN PROGRESS***

@@

\- Notes (Will be expanded and explained in next update):

-   Accuracy: 85.09%

-   Target variable may not be balanced

@@

\- Notes (Will be expanded and explained in next update):

-   Model is only predicting '0', likely an error in model setup

-   Due to inaccurate predictions, accuracy is not applicable.

@@

-   in progress

@@

## Full analysis

***\*To be updated in later portion of project***

@@

***\*To be updated in later portion of project***

@@

***\*To be updated in later portion of project***

-   LIMIT: Having a lot of categorical features

-   LIMIT: Having a response variable with 3 levels instead of a binary.

@@

***\*To be updated in later portion of project***

@@

***\*To be updated in later portion of project***

>>>>

---
title: "Manuscript for a Data Analysis Project: Diabetes Diagnosis"
subtitle: ""
author: Austin Thrash
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

@@

Diabetes is a chronic condition that affects millions of people worldwide, characterized by high blood glucose levels and serious complications if not managed properly. This study aims to identify key health indicators predicting the likelihood of an individual being diagnosed with diabetes using a dataset from the UC Irvine Machine Learning Repository, containing over 250,000 instances and 21 features. Three models were utilized: logistic regression, random forest, and K-nearest neighbors (KNN). The analysis revealed that BMI, high blood pressure, heart disease or attack history, and heavy alcohol consumption were significant predictors of diabetes. Logistic regression and KNN models demonstrated reasonable accuracy, while the random forest model required refinement. The study highlights the importance of these health indicators and suggests further data processing and model tuning for improved predictive performance.

@@

"Diabetes can lead to severe, and in some cases, life-threatening complications." [@ADA]. Diabetes is a major cause of blindness, kidney failure, heart attacks, stroke, and lower limb amputation. Managing diabetes and its complications places a significant burden on healthcare systems worldwide. Early diagnosis and management are crucial to prevent adverse health outcomes. Several studies have identified various risk factors associated with diabetes, including age, BMI, physical activity, and dietary habits.

@@

|---------------------|---------------------------------------------------|

@@

For my model selection, I referenced ISLR2. [@ISLR]

@@

```{r}
knitr::include_graphics(here("results","figures","importance_table.png"))
```

@@

@@

The logistic regression model was applied to the dataset to analyze the relationship between the binary outcome variable (diabetes diagnosis) and the predictors. The model achieved an accuracy of 85.09%. However, the model showed that the target variable might not be balanced, indicating that further data processing might be necessary.

@@

The random forest model was applied to the dataset to improve the accuracy and robustness of predictions. The model, however, only predicted the class '0', likely due to an error in model setup. This resulted in an accuracy that was not applicable, indicating the need for further refinement of the model.

@@

          Reference
Prediction    X0    X1    X2
        X0 21270   415  2975
        X1     0     0     0
        X2   285    16   539

               Accuracy : 0.8553          
                 95% CI : (0.8509, 0.8596)
    No Information Rate : 0.8453          
    P-Value [Acc > NIR] : 4.769e-06

                  Kappa : 0.1869

                     Class: X0 Class: X1 Class: X2
Sensitivity             0.9868    0.0000   0.15339
Specificity             0.1407    1.0000   0.98631
Pos Pred Value          0.8625       NaN   0.64167
Neg Pred Value          0.6607    0.9831   0.87936
Prevalence              0.8453    0.0169   0.13780
Detection Rate          0.8341    0.0000   0.02114
Detection Prevalence    0.9671    0.0000   0.03294
Balanced Accuracy       0.5637    0.5000   0.56985
```

The KNN model was applied to the dataset and achieved an accuracy of 85.53%. The model's sensitivity and specificity values indicated that it performed reasonably well, but there was room for improvement, particularly in predicting minority classes.

@@

@@

The analysis revealed that BMI, HighBP, HeartDiseaseorAttack, and HvyAlchoholConsump were significant predictors of diabetes diagnosis. Logistic regression and KNN models showed reasonable accuracy, while the random forest model required further refinement. These findings highlight the importance of these health indicators in predicting diabetes and suggest that further data processing and model tuning could improve predictive performance.

@@

-   **Imbalanced Dataset**: The dataset was imbalanced, which affected the model's performance, particularly in predicting minority classes.

-   **Categorical Data Handling**: The handling of categorical variables posed challenges, especially for models like logistic regression and random forest.

-   **Model Setup Errors**: The random forest model's performance indicated potential errors in model setup, necessitating further refinement.

@@

The study identified key health indicators that are strongly associated with diabetes diagnosis. Logistic regression and KNN models provided reasonable predictive accuracy, while the random forest model required further refinement. These findings underscore the need for balanced datasets and appropriate handling of categorical variables to improve model performance.

@@

For some reason my bib file was being weird, will try to fix.
